### 2017-09-21 Article analysis.Rmd
Beginnings

---
title: "Article analysis"
author: "Maria C. Codlin"
date: "September 21, 2017"
output: html_document
---
Before I run any parts of the analysis, I load all the packages I use in this file
```{r load libraries}
require(dplyr)
require(ggplot2)
```
First in excel, I deleted two rows of non-table data, i.e. title, and a note, and saved the file (S2_Table) as a CSV in the project directory. Then I read it into R and converted it to a table using dplyr
```{r readfile}
iso<-read.csv("S2_Table.csv")
iso<-tbl_df(iso)
head(iso)
```

To make life easier, I changed the names of two key variables. First I tried 18O and 13C as the variable names, but it through me an error, I think by starting with a number. 
```{r }
names(iso)[12]<-"C13ap"
names(iso)[13]<-"O18ap"
names(iso)
```
This is my first attempt at a box plot! I figured it out eventually
```{r }
g<-ggplot(iso, aes(x=Site, y=O18ap))
g+geom_boxplot(aes(color=Genus))
```

```{r }

```

### 2017-09-21 Article analysis.Rmd
workin
working

@@
 g<-ggplot(iso, aes(x=Site, y=O18ap))
-g+geom_boxplot(aes(color=Genus))
+g+geom_boxplot(aes(color=Genus))+labs(x="" ,y="δ18O apatite-VSMOW (‰)")
+```
+Continuation of cleaning.
+Some samples were removed from analysis due to the likely hood of contamination or diagenesis.
+" Five apatite samples with C/P ratios <0.10, three apatite samples with IR-SF values < 2.00, and five apatite samples with IR-SF values >4.00 were excluded from subsequent discussions"
+
+This code deleted all my data?
+```{r }
+iso_total<-iso
 ```
@@
 ```{r }
+iso<-filter(iso_total, C.P>=0.10, IR.SF>=2.00, IR.SF<=4.00)
 
+iso
 ```

### 2017-09-24 Article analysis notebook.Rmd
COMMITTMENT

---
title: "Article Analysis Notebook"
author: Maria C. Codlin
output: html_notebook
---
###Table of contents?

###Introduction to this document, the contents, and the article.

To start with a blank slate
```{r clear workspace}
rm(list=ls())
```

#Step 1. Load packages and data
Before I run any parts of the analysis, I load all the packages I use in this file
```{r load libraries}
require(dplyr)
require(ggplot2)
require(stringr)
require(plotrix) #to calculate standard error
```
First in excel, I deleted two rows of non-table data, i.e. title, and a note, and saved the file (S2_Table) as a CSV in the project directory. Then I read it into R and converted it to a table using dplyr. I trimed white space at beginnings and ends of things where there shouldn't be any, and kept strings as is to avoid issues later down the line. 
```{r readfile arch}
iso_total<-read.csv("S2_Table.csv", header=TRUE, stringsAsFactors = FALSE, strip.white=TRUE)
iso_total<-tbl_df(iso_total)
head(iso_total)
str(iso_total)
```
```{r readfile modern}
modern<-read.csv("S1_Table.csv", header=TRUE, skip=3, strip.white=TRUE, stringsAsFactors = FALSE) #here I used skip=3 to skip the first 3 rows before the data started
modern<-tbl_df(modern)
head(modern)
str(modern)
```

To make life easier, I changed the names of two key variables. First I tried 18O and 13C as the variable names, but it threw me an error, I think by starting with a number. 
```{r rename}
names(iso_total)[12]<-"C13ap" #I used names(iso_total) to get the position of these for renaming
names(iso_total)[13]<-"O18ap"
names(modern)[10]<-"modC13"
names(modern)[11]<-"modO18"
```

#Step 2: Remove of samples potentially affected by diagenisis and summarize data to check against means presented in the article

In this article, they tested a random sample (n=101) of bone for potential post-depositional alteration to the mineral structure.Using FTIR, they tested the carbonate to phospate ratios (C/P) and infrared splitting factor (IR-SF) using FTIR and removed samples which fell outside of the expected ranges of 0.1 and 0.5 for C/P and 2.0 and 4.0 for IR-SF. The samples that they removed are marked in grey in the supplimentary data.

Here, I used the filter function to select only those rows where C/P and IR-SF fell within the expected ranges, and saved this as a new dataset "iso". I also needed to make sure the rows with NA values were still included, since only a random sample was analysed. I did this using the | notation which functions like "OR". Since the FTIR analysis returned both C/P and IR-SF values, I only need to include one is.na function here. I also excluded an outlier with an extremely low delta O18 that they excluded by converting this number to NA, as the samples delta C13 value was still incorperated in the analysis.
```{r filter out diagenisis}
#to filter out only Lepus and Sylvilagus genus, I first needed to define the column as a factor
iso_total$Genus<-as.factor(iso_total$Genus)
iso<-filter(iso_total, (Genus=="Lepus" | Genus=="Sylvilagus") & between (C.P, 0.1, 0.5)&between (IR.SF, 2.0, 4.0) | (is.na(IR.SF))) #brackets around the Genus filter were important here, otherwise the | worked on everything that came after
iso$O18ap[iso$O18ap==11.9]<-NA #couldn't figure out how to get this to work with dplyr or chaining
iso            
```

"δ13Capatite = -7.8 ± 2.4‰(N = 114, 1 S.D.) and δ18Oapatite = 26.0‰(N = 113, 1 S.D.). The modern reference specimens exhibited
mean values of δ13Capatite = -12.4 ± 1.8‰(N = 13, 1 S.D.) and δ18Oapatite = 26.1‰± 1.4 (N = 13, 1 S.D.)."(Somerville et al. 2016:10). *I also note that they should have used "n" here*

Note: the first time I tried this it was not a match and so I realised I needed to keep the na values in the filter above. The second time it was also not correct so I added two lines to count the number of values in the two columns of interest to see if the sample sizes match. I noted the single value from delta O18 to remove but I still have one extra row in there than their sample number. I checked on the excel, and there is no evidence of this extra row that they excluded. I took a guess that this may have been a species listed in the spreadsheet as Lepus? and removal of this specimen now produces averages that match the data presented in the article.

```{r sum_table}
my_sum <- 
  list("Archaeological samples" = 
         list("δC13 mean" = ~ mean(C13ap, na.rm=TRUE),
              "δC13 sd" = ~ sd(C13ap, na.rm=TRUE),
              "δO18 mean" = ~ mean(O18ap, na.rm=TRUE),
              "δO18 sd" = ~ sd(O18ap, na.rm=TRUE)),
       "Modern samples" = 
         list("δC13 mean" = ~ mean(modC13, na.rm=TRUE),
              "δC13 sd" = ~ sd(modC13, na.rm=TRUE),
              "δO18 mean" = ~ mean(modO18, na.rm=TRUE),
              "δO18 sd" = ~ sd(modO18, na.rm=TRUE))
  )

summary_table(iso, my_sum[1])
summary_table(modern, my_sum[2])
```

```{r summary}
sum(!is.na(iso$C13ap))#the ! means it will return the number that is not NA
sum(!is.na(iso$O18ap))
arch_sum<-summarize(iso, meanC13=mean(C13ap, na.rm=TRUE), meanO18=mean(O18ap, na.rm=TRUE), sdC13=sd(C13ap, na.rm=TRUE), sdO18=sd(O18ap, na.rm=TRUE))%>%print()

mod_sum<-summarize(modern, meanC13=mean(modC13, na.rm=TRUE), meanO18=mean(modO18, na.rm=TRUE), seC13=std.error(modC13, na.rm=TRUE), sdO18=sd(modO18, na.rm=TRUE))%>%print()

```

##Step 3

### 2017-09-24 summary_table.R

@@ -1,0 +1,17 @@
+#Building a summary list for a summary table
+
+my_sum <- 
+  list("Archaeological samples" = 
+         list("δC13 mean" = ~ mean(C13ap, na.rm=TRUE),
+              "δC13 sd" = ~ sd(C13ap, na.rm=TRUE),
+              "δO18 mean" = ~ mean(O18ap, na.rm=TRUE),
+              "δO18 sd" = ~ sd(O18ap, na.rm=TRUE)),
+       "Modern samples" = 
+         list("δC13 mean" = ~ mean(modC13, na.rm=TRUE),
+              "δC13 sd" = ~ sd(modC13, na.rm=TRUE),
+              "δO18 mean" = ~ mean(modO18, na.rm=TRUE),
+              "δO18 sd" = ~ sd(modO18, na.rm=TRUE))
+  )
+
+summary_table(iso, my_sum[1])
+summary_table(modern, my_sum[2])

### 2017-09-27 Article analysis.Rmd

@@
 ---
+###Table of contents?
+
+###Introduction to this document, the contents, and the article.
+
 Before I run any parts of the analysis, I load all the packages I use in this file
@@
 iso<-read.csv("S2_Table.csv")
-iso<-tbl_df(iso)
-head(iso)
+iso<-tbl_df(iso_total)
+head(iso_total)
 ```
 
-To make life easier, I changed the names of two key variables. First I tried 18O and 13C as the variable names, but it through me an error, I think by starting with a number. 
+To make life easier, I changed the names of two key variables. First I tried 18O and 13C as the variable names, but it threw me an error, I think by starting with a number. 
 ```{r }
@@
 
-This code deleted all my data?
 ```{r }
-iso_total<-iso
+iso<-filter(iso_total, between (C.P, 0.1, 0.5)&between (IR.SF, 2.0, 4.0))
+iso            
 ```
 
-```{r }
-iso<-filter(iso_total, C.P>=0.10, IR.SF>=2.00, IR.SF<=4.00)
+###Outline of what needs to be done. 
+Datasets: 1. Isotopes of modern specimens. 2. Isotopes of ancient specimens.
 
-iso
+##Modification of dataset prior to analysis
+- calibration of carbon isotope ratios due to atmospheric changes in the past 130 years to make the modern specimens in the dataset comparable to the ancient ones.**It is not clear whether this is incorperated in the supplimental data or not**
+
+##Statistical analyses:
+- Parametric tests: Independent samples t-test (groups of two) and ANOVA (and two or more)
+-Non parametic: Mann-Whitney U test (for Oztoyahualco and pooled other areas) due to very different sample sizes.
+-To see if carbon and oxygen isotopes covary, bivariate Pearson's correlation was used. Sig assumed at a=0.05.
+
+##Mixing model to approximate percentage of C4 and CAM plants in diet:
+%C4CAM= ((delta_13_C_apatite - 9.7)-(-26.1))/14.1
+Due to the level of uncertainties, these percentages are used to reflect the relative contrabutions of C4 and CAM plants between individuals rather than as absolute data.
+
+```{r summary}
+summarise(iso)
+
 ```
+
+*Because of errors later on, I found I also needed to trim the space at the end of some of the Genus names.* 
+```{r trim white space}
+iso_total$Genus<-str_trim(iso_total$Genus)
+iso_total$Genus<-as.factor(iso_total$Genus)
+levels(iso_total$Genus)
+```
+
+```{r sum_table}
+my_sum <- 
+  list("Archaeological samples" = 
+         list("δC13 mean (sd)" = ~ mean_sd(C13ap),
+              "δO18 mean (sd)" = ~ mean_sd(O18ap)),
+       "Modern samples" = 
+         list("δC13 mean (sd)" = ~ mean_sd(modern$modC13),
+              "δO18 mean (sd)" = ~ mean_sd(modern$modO18))
+              )
+
+summary_table(iso, my_sum[1])
+summary_table(iso, my_sum[2])
+```

### 2017-10-03 Article analysis.Rmd
librarys

@@
 require(dplyr)
+require(stringr)
 require(ggplot2)
+require(qwraps2)
 ```

### 2017-10-07 Article analysis notebook.Rmd
Friday - error in dataset

@@
 require(plotrix) #to calculate standard error
+require(qwraps2)#for summary table
+require(tidyr)
 ```
@@
 ```{r readfile arch}
-iso_total<-read.csv("S2_Table.csv", header=TRUE, stringsAsFactors = FALSE, strip.white=TRUE)
+iso_total<-read.csv("S2_Table.csv", header=TRUE, stringsAsFactors = FALSE, strip.white=TRUE, na.strings = c("", "NA")) # there were some rows of blanks being included so I used this last bit to convert them to NA
 iso_total<-tbl_df(iso_total)
@@
 ```{r readfile modern}
-modern<-read.csv("S1_Table.csv", header=TRUE, skip=3, strip.white=TRUE, stringsAsFactors = FALSE) #here I used skip=3 to skip the first 3 rows before the data started
+modern<-read.csv("S1_Table.csv", header=TRUE, skip=3, strip.white=TRUE, stringsAsFactors = FALSE, na.strings = c("", "NA")) #here I used skip=3 to skip the first 3 rows before the data started
 modern<-tbl_df(modern)
@@
 names(iso_total)[13]<-"O18ap"
+names(iso_total)[1]<-"lab_n"
 names(modern)[10]<-"modC13"
@@
 ```
+```{r}
+iso<-filter(iso, !(is.na(iso$Genus))) #removes empty rows
+```
 
-"δ13Capatite = -7.8 ± 2.4‰(N = 114, 1 S.D.) and δ18Oapatite = 26.0‰(N = 113, 1 S.D.). The modern reference specimens exhibited
-mean values of δ13Capatite = -12.4 ± 1.8‰(N = 13, 1 S.D.) and δ18Oapatite = 26.1‰± 1.4 (N = 13, 1 S.D.)."(Somerville et al. 2016:10). *I also note that they should have used "n" here*
+**Descriptive statistics from the article**
+Archaological specimens: δ13Capatite = -7.8 ± 2.4‰(N = 114, 1 S.D.) and δ18Oapatite = 26.0‰(N = 113, 1 S.D.). 
+Modern reference specimens: δ13Capatite = -12.4 ± 1.8‰(N = 13, 1 S.D.) and δ18Oapatite = 26.1‰± 1.4 (N = 13, 1 S.D.)."(Somerville et al. 2016:10). *I also note that they should have used "n" here*
 
@@
   list("Archaeological samples" = 
-         list("δC13 mean" = ~ mean(C13ap, na.rm=TRUE),
-              "δC13 sd" = ~ sd(C13ap, na.rm=TRUE),
-              "δO18 mean" = ~ mean(O18ap, na.rm=TRUE),
-              "δO18 sd" = ~ sd(O18ap, na.rm=TRUE)),
+        round (c("δC13 n" = sum(!is.na(iso$C13ap)),
+              "δC13 mean" =  mean(iso$C13ap, na.rm=TRUE),
+              "δC13 sd" =  sd(iso$C13ap, na.rm=TRUE),
+              "δO18 n" = sum(!is.na(iso$O18ap)),
+              "δO18 mean" =  mean(iso$O18ap, na.rm=TRUE),
+              "δO18 sd" =  sd(iso$O18ap, na.rm=TRUE)),2),
        "Modern samples" = 
-         list("δC13 mean" = ~ mean(modC13, na.rm=TRUE),
-              "δC13 sd" = ~ sd(modC13, na.rm=TRUE),
-              "δO18 mean" = ~ mean(modO18, na.rm=TRUE),
-              "δO18 sd" = ~ sd(modO18, na.rm=TRUE))
+         round(c("δC13 n" = sum(!is.na(modern$modC13)),
+              "δC13 mean" =  mean(modern$modC13, na.rm=TRUE),
+              "δC13 sd" =  sd(modern$modC13, na.rm=TRUE),
+              "δO18 n" = sum(!is.na(modern$modO18)),
+              "δO18 mean" =  mean(modern$modO18, na.rm=TRUE),
+              "δO18 sd" =  sd(modern$modO18, na.rm=TRUE)),2)
        )
 
-summary_table(iso, my_sum[1])
-summary_table(modern, my_sum[2])
+my_sum
 ```
 
-```{r summary}
-sum(!is.na(iso$C13ap))#the ! means it will return the number that is not NA
-sum(!is.na(iso$O18ap))
-arch_sum<-summarize(iso, meanC13=mean(C13ap, na.rm=TRUE), meanO18=mean(O18ap, na.rm=TRUE), sdC13=sd(C13ap, na.rm=TRUE), sdO18=sd(O18ap, na.rm=TRUE))%>%print()
+#Step 3 "Independent samples t-tests were conducted to compare mean stable isotope ratios of the two genera (Lepus and Sylvilagus) across the archaeological bone sample as a whole.
+No significant difference between the genera were found in δ13Capatite values (t[110] = -0.806, p = 0.422) or δ18Oapatite values (t[110] = 0.428, p = 0.669)."
 
-mod_sum<-summarize(modern, meanC13=mean(modC13, na.rm=TRUE), meanO18=mean(modO18, na.rm=TRUE), seC13=std.error(modC13, na.rm=TRUE), sdO18=sd(modO18, na.rm=TRUE))%>%print()
+First, I will check that the data fits a normal distribution by comparing it to the qnorm and qqline plots as well as histogram.
 
+**Carbon Apatite Results**
+*Histogram*
+```{r}
+a_hist<-ggplot(iso, aes(C13ap))
+a_hist+geom_histogram(bins= 15)+ facet_wrap(~Genus, scales="free")
+```
+Sylvilagus looks approximately normal while Lepus it is hard to tell. Next I compare using qqplots.
+*QQplots*
+```{r qqplots}
+LC<-iso$C13ap[iso$Genus=="Lepus"]
+SC<-iso$C13ap[iso$Genus=="Sylvilagus"]
+qqnorm(LC,xlab="Theoretical",main="Lepus")
+  qqline(LC)
+
+  qqnorm(SC,xlab="Theoretical",main="Sylvilagus")
+  qqline(SC)
+```
+Again, Lepus fits this "OK" while Sylvilagus looks pretty good. 
+*Test of equal variance*
+```{r}
+var.test(LC, SC)
 ```
 
-##Step 3
+This suggests that the variance of the two samples is comparable and thus a two sample T test can be used.First I report the tests for the oxygen data and then provide both t-tests together with a discussion.
+
+**Oxygen Apatite Results**
+*Histogram*
+```{r}
+a_hist<-ggplot(iso, aes(O18ap))
+a_hist+geom_histogram(bins=15)+ facet_wrap(~Genus, scales="free")
+```
+Both look like they could be approximately normal
+*QQplots*
+```{r qqplots}
+LO<-iso$O18ap[iso$Genus=="Lepus"]
+SO<-iso$O18ap[iso$Genus=="Sylvilagus"]
+qqnorm(LO,xlab="Theoretical",main="Lepus")
+  qqline(LO)
+
+  qqnorm(SO,xlab="Theoretical",main="Sylvilagus")
+  qqline(SO)
+```
+Here, Lepus fits the normal distrbution well, while Sylvilagus is divergent. Looking at the histgram again, it could poentitally be bimodal. 
+*Test of equal variance*
+```{r}
+var.test(LO, SO)
+```
+*Carbon apatite two sample T-test*
+```{r}
+t.test(LC, SC, var.equal =TRUE, alternative = "two.sided")
+```
+
+*Oxygen apatite two sample T-test*
+```{r}
+t.test(SO, LO, var.equal =TRUE, alternative = "two.sided")
+```
+*Summary:*
+Carbon apatite t-sample t-test comparing Lepus and Sylvilagus: t = 0.5185, df = 112, p-value = 0.6051
+Oxygen apatite t-sample t-test comparing Lepus and Sylvilagus: t = -0.83133, df = 111, p-value = 0.4076
+*Values reported in Somerville et al. (2016:10)*
+δ13Capatite values (t[110] = -0.806, p = 0.422) 
+δ18Oapatite values (t[110] = 0.428, p = 0.669).
+
+Both of these tests returned different test statistics, degrees of freedom and p-values to those reported by Somerville et al.
+(2016).
+I tried multiple variations of the t-test in R, from Welsh t-test and one sided t-tests to one sided Welsh t-tests and none of the tests produced a similar result to those reported by Somerville.
+In fact, the two sample t-test for equal varience that I ran for both carbon and oxygen data matches most closely with the result reported for the opposite isotopic data.
+i.e.
+my oxygen looks most similar to their carbon.
+I have a number of hypothesis for why these tests may vary: the SPSS test is vastly different to the R t-tests, they are excluding samples from the test and not reporting it, as indicated by the difference in df between our tests, or there was an error in their reporting of the data.
+That said, the outcome of all tests is the same, there is no significant difference in the means of the samples.
+I am also not sure about the utility of this test in general, given that the article is looking for differences within a single species as well to test for outliers in the diet of these animals.
+
+#Step 4 
+"We also evaluated stable isotope ratios between the genera and between individual site locations using a two-way ANOVA (factorial ANOVA ith two factors), testing for the main effects of genus and context, and for the genuscontext interaction."
+
+```{r}
+carbon<-lm(iso$C13ap~iso$Genus*iso$Site)
+anova(carbon)
+```
+```{r}
+oxygen<-lm(iso$O18ap~iso$Genus*iso$Site)
+anova(oxygen)
+```

### 2017-10-07 Article analysis.Rmd
Friday - error in dataset

@@
 ```
+
+```{r summary}
+sum(!is.na(iso$C13ap))#the ! means it will return the number that is not NA
+sum(!is.na(iso$O18ap))
+arch_sum<-summarize(iso, meanC13=mean(C13ap, na.rm=TRUE), meanO18=mean(O18ap, na.rm=TRUE), sdC13=sd(C13ap, na.rm=TRUE), sdO18=sd(O18ap, na.rm=TRUE))%>%print()
+
+mod_sum<-summarize(modern, meanC13=mean(modC13, na.rm=TRUE), meanO18=mean(modO18, na.rm=TRUE), seC13=std.error(modC13, na.rm=TRUE), sdO18=sd(modO18, na.rm=TRUE))%>%print()
+
+```
+
+These results also differ from what was presented. So I will test to see if they removed outliers prior to t-test. I will use a box plot to identify outliers
+```{r}
+boxplot(SC, ylim = c(), main = "dC13 ap", xlab = "Sylvilagus")
+boxplot(LC, ylim = c(), main = "dC13 ap", xlab = "Lepus")
+boxplot(SO, ylim = c(), main = "dO18 ap", xlab = "Sylvilagus")
+boxplot(LO, ylim = c(), main = "dO18 ap", xlab = "Lepus")
+```

### 2017-10-07 summary stats 2.Rmd
Friday - error in dataset

```{r sum_table}
my_sum <- 
  list("Lepus" = 
        round (c("δC13 n" = sum(!is.na(iso$C13ap)),
              "δC13 mean" =  mean(iso$C13ap, na.rm=TRUE),
              "δC13 sd" =  sd(iso$C13ap, na.rm=TRUE),
              "δO18 n" = sum(!is.na(iso$O18ap)),
              "δO18 mean" =  mean(iso$O18ap, na.rm=TRUE),
              "δO18 sd" =  sd(iso$O18ap, na.rm=TRUE)),2),
       "Sylvilagus" = 
         round(c("δC13 n" = sum(!is.na(modern$modC13)),
              "δC13 mean" =  mean(modern$modC13, na.rm=TRUE),
              "δC13 sd" =  sd(modern$modC13, na.rm=TRUE),
              "δO18 n" = sum(!is.na(modern$modO18)),
              "δO18 mean" =  mean(modern$modO18, na.rm=TRUE),
              "δO18 sd" =  sd(modern$modO18, na.rm=TRUE)),2)
       )

my_sum
```

```{r}
sum2<-function(x, y){
for(i in levels(iso$Site)){
  for(i in levels(iso$Genus)){
  x1 <- with(iso$Genus, x[Genus==i])
  y1 <- with(iso$Genus, y[Genus==i])
result<- summarise(mean (x1, na.rm=TRUE))

  }
  return(result)}
}

```
```{r}
sum2(x=iso$C13ap, y=iso$O18ap)
```

round (c("δC13 n" = sum(!is.na(x1)),
              "δC13 mean" =  mean(x1, na.rm=TRUE),
              "δC13 sd" =  sd(x1, na.rm=TRUE),
              "δO18 n" = sum(!is.na(y1)),
              "δO18 mean" =  mean(y1, na.rm=TRUE),
              "δO18 sd" =  sd(y1, na.rm=TRUE)),2)

```{r}
library(dplyr)
library(tidyr)
iso %>%
  group_by(Genus, Site) %>%
  summarize(meanC13=mean(C13ap, na.rm=TRUE)) %>%
  spread(Genus, meanC13)
```
```{r}
carbon_group<-iso %>%
  select(lab_n:Genus, C13ap)%>%
  filter(!(is.na(C13ap)))%>%
  group_by(Site, Genus)
carbon_group
oxygen_group<-iso %>%
  select(lab_n:Genus, O18ap)%>%
  filter(!(is.na(O18ap)))%>%
  group_by(Site, Genus)

carbon_group
oxygen_group
```
```{R}
summ<-carbon_group%>%
  summarize(length(C13ap), mean(C13ap), sd(C13ap))
summ2<-oxygen_group %>%
  summarize(length(O18ap), mean(O18ap), sd(O18ap))

M<-merge(summ,summ2)
M[,3:8]<-round(M[3:8], 2)
M
sum(M$`length(C13ap)`)

```
round(M, 2)
```
t1<-iso$C13ap %>%
  group_by(Site, Genus) %>%
  count()
t2<-iso$O18ap %>%
  group_by(Site, Genus) %>% 
  length()
summ
t1
m1<-cbind(summ, t1)
m2<-cbind(summ2, t2)
m1
merge(m1, m2)
```

```{r}
m<-NULL

iso %>%

 for(i in levels(iso$Genus)){
   tmp <- with(iso, C13ap[Genus=="Lepus"])
        m$δC13_n=sum(!is.na(tmp))

       }

```

### 2017-10-26 Article analysis notebook.Rmd
where did the summary go?

--- 2017-09-24 Article analysis notebook.Rmd
@@
 ###Introduction to this document, the contents, and the article.
+
+First I modified the files slightly to import with curl, as it refused to bring in the header names with certain symbols in (it worked fine importing from local repo).
 
@@
 require(plotrix) #to calculate standard error
+require(qwraps2)
 ```
@@
 ```{r readfile arch}
-iso_total<-read.csv("S2_Table.csv", header=TRUE, stringsAsFactors = FALSE, strip.white=TRUE)
+library(curl)
+f <- curl("https://raw.githubusercontent.com/mazattack/Article-analysis/master/S2_Table.csv")
+iso_total <- read.csv(f, header = TRUE, stringsAsFactors = FALSE, strip.white=TRUE)
 iso_total<-tbl_df(iso_total)
@@
 ```{r readfile modern}
-modern<-read.csv("S1_Table.csv", header=TRUE, skip=3, strip.white=TRUE, stringsAsFactors = FALSE) #here I used skip=3 to skip the first 3 rows before the data started
+f <- curl("https://raw.githubusercontent.com/mazattack/Article-analysis/master/S1_Table.csv")
+modern<-read.csv(f, header=TRUE, sep = ",", skip=0, strip.white=TRUE, stringsAsFactors = FALSE) #here I used skip=3 to skip the first 3 rows before the data started
 modern<-tbl_df(modern)

### 2017-10-27 Article analysis notebook.Rmd
mac
-check for updated vesion on windows

--- 2017-10-07 Article analysis notebook.Rmd
@@
 title: "Article Analysis Notebook"
-author: Maria C. Codlin
-output: html_notebook
+author: "Maria C. Codlin"
+output:
+  pdf_document: default
+  html_notebook: default
 ---
@@
 require(plotrix) #to calculate standard error
-require(qwraps2)#for summary table
+require(qwraps2)
 require(tidyr)
@@
 ```{r readfile arch}
-iso_total<-read.csv("S2_Table.csv", header=TRUE, stringsAsFactors = FALSE, strip.white=TRUE, na.strings = c("", "NA")) # there were some rows of blanks being included so I used this last bit to convert them to NA
+library(curl)
+f <- curl("https://raw.githubusercontent.com/mazattack/Article-analysis/master/S2_Table.csv")
+iso_total <- read.csv(f, header = TRUE, stringsAsFactors = FALSE, strip.white=TRUE)
+```
+
 iso_total<-tbl_df(iso_total)
@@
 ```{r readfile modern}
-modern<-read.csv("S1_Table.csv", header=TRUE, skip=3, strip.white=TRUE, stringsAsFactors = FALSE, na.strings = c("", "NA")) #here I used skip=3 to skip the first 3 rows before the data started
+
+f <- curl("https://raw.githubusercontent.com/mazattack/Article-analysis/master/S1_Table.csv")
+modern<-read.csv(f, header=TRUE, sep = ",", skip=0, strip.white=TRUE, stringsAsFactors = FALSE) #here I used skip=3 to skip the first 3 rows before the data started
+
 modern<-tbl_df(modern)
@@
 ```{r}
+stargazer(attitude, header=FALSE)
+```
+```{r}
 carbon<-lm(iso$C13ap~iso$Genus*iso$Site)

### 2017-10-29 Article analysis notebook.Rmd
ANOVA missing pyramid

@@
 output:
+  html_notebook: default
+  html_document: default
   pdf_document: default
-  html_notebook: default
 ---
@@
 require(tidyr)
+require(stargazer)
+require(knitr)
 ```
@@
 ```
-
+```{r}
 iso_total<-tbl_df(iso_total)
@@
 ```
+
 ```{r readfile modern}
@@
 *QQplots*
-```{r qqplots}
+```{r qqplots carbon}
 LC<-iso$C13ap[iso$Genus=="Lepus"]
@@
   qqline(SC)
+
+  shapiro.test(LC)
+  shapiro.test(SC)
 ```
-Again, Lepus fits this "OK" while Sylvilagus looks pretty good. 
+Again, Lepus fits this "OK" while Sylvilagus looks pretty good. Both samples pass the Shapiro-Wilk test for normality. 
 *Test of equal variance*
@@
 *QQplots*
-```{r qqplots}
+```{r qqplots Oxygen}
 LO<-iso$O18ap[iso$Genus=="Lepus"]
@@
 That said, the outcome of all tests is the same, there is no significant difference in the means of the samples.
-I am also not sure about the utility of this test in general, given that the article is looking for differences within a single species as well to test for outliers in the diet of these animals.
+I later realised that the data they provide is slightly differerent from the results presented and this is likely the reason for the differences observed in this test.
+This is most clearly observed in the table below and comparison to their table 2.
 
 #Step 4 
-"We also evaluated stable isotope ratios between the genera and between individual site locations using a two-way ANOVA (factorial ANOVA ith two factors), testing for the main effects of genus and context, and for the genuscontext interaction."
+"We also evaluated stable isotope ratios between the genera and between individual site locations using a two-way ANOVA (factorial ANOVA with two factors), testing for the main effects of genus and context, and for the genuscontext interaction."
+
+*Summary table (Compare Table 2)*
+
+```{r create genus column in modern}
+Genus<- as.factor(modern$Species)
+modern<-cbind(modern, Genus)
+levels(modern$Genus)[levels(modern$Genus)=="Lepus californicus texianus"]<-"Lepus"
+levels(modern$Genus)[levels(modern$Genus)=="Sylvilagus floridanus"]<-"Sylvilagus"
+levels(modern$Genus)[levels(modern$Genus)=="Sylvilagus floridanus orizabae"]<-"Sylvilagus"
+levels(modern$Genus)    
+```
+
+```{r merge modern and iso datasets}
+new<-select(iso, lab_n, Site, Phase, Genus, C13ap, O18ap)
+new2<-select(modern, lab_n=Lab_num, Site=State, Phase=Collection.date..yrs.AD., Genus, C13ap=modC13, O18ap=modO18)%>%mutate(Site="modern")
+iso2<-rbind(filter(new,(!(is.na(C13ap)))), new2)
+iso2
+```
+```{r constructing the summary table}
+iso2$Genus <-as.factor(iso2$Genus)
+iso2_g <- group_by(iso2, Site, Genus)
+sum_tot<- summarize(iso2_g, length(C13ap), mean(C13ap), sd(C13ap), length(O18ap), mean(O18ap), sd(O18ap))
+
+tot_tot<-iso2%>%
+  group_by(Genus)%>%
+  summarize(length(C13ap), mean(C13ap), sd(C13ap), length(O18ap), mean(O18ap), sd(O18ap))
+tot_tot<-mutate(tot_tot, Site="Total")
+tot<-rbind.data.frame(sum_tot, tot_tot)
+tot
+```
+Note the differences across these summary statistics compared to the table provided in the article (Table 2).
+This reflects a difference in the data from the article compared to what they provided.
+I checked this in excel to confirm.
 
 ```{r}
-stargazer(attitude, header=FALSE)
-```
-```{r}
 carbon<-lm(iso$C13ap~iso$Genus*iso$Site)
-anova(carbon)
+anova(carbon) #not sure why but moon pyramid is missing from all these ANOVA
 ```
@@
 oxygen<-lm(iso$O18ap~iso$Genus*iso$Site)
-anova(oxygen)
+summary(oxygen)
 ```
+My two way Anova summaries match the conclusions they derived in the article.
+That Genus had no significant effect on either δ13C or δO18 apatite ratios.
+However there is a significant effect of Site location on these variables.
+
+**One way ANOVA tests**
+```{r}
+carbon<-lm(iso$C13ap~iso$Site)
+summary(carbon)
+```
+
+```{r}
+oxygen<-lm(iso$O18ap~iso$Site)
+summary(oxygen)
+```
+
+```{r}
+
+```
+
+, the main effect for genus was not significant (F[1,102] =
+3.361), p = 0.070), but the main effect of site context was significant (F[4,102] = 5.365,
+p = 0.001). Importantly, the interaction between genus and site was not significant for δ13Capatite
+(F[4,102] = 0.590, p<0.670) or for δ18Oapatite (F[4,102] = 0.834, p<0.507). However, Lepus and
+Sylvilagus values are notably different from each other at the Cuevas site, but the small sample
+size from this context makes any interpretations regarding this observation speculative. Subsequent
+contextual comparisons thus combine Sylvilagus and Lepus specimens
+
+*For δ13Capatite, the main effect for genus was not significant (F[1,102] =
+3.361), p = 0.070), but the main effect of site context was significant (F[4,102] = 5.365,
+p = 0.001). Importantly, the interaction between genus and site was not significant for δ13Capatite
+(F[4,102] = 0.590, p<0.670) or for δ18Oapatite (F[4,102] = 0.834, p<0.507). 
+However, Lepus and
+Sylvilagus values are notably different from each other at the Cuevas site, but the small sample
+size from this context makes any interpretations regarding this observation speculative. Subsequent
+contextual comparisons thus combine Sylvilagus and Lepus specimens*

### 2017-10-29 Article analysis.Rmd
mac
-check for updated vesion on windows
Mac (#1)
* where did the summary go?

* mac

-check for updated vesion on windows

* selet
ANOVA missing pyramid

@@
 
+```{r}
+stargazer(iso_total$Element, align = TRUE)
+
+```
+
 ```{r summary}
@@
 ```
+
+*Working out the summary table*
+```{r group by genus}
+modern_group<-modern %>%
+  select(modC13:Genus)%>%
+  group_by(Genus)
+```
+
+```{R}
+summ<-carbon_group%>%
+  summarize(length(C13ap), mean(C13ap), sd(C13ap))
+summ2<-oxygen_group %>%
+  summarize(length(O18ap), mean(O18ap), sd(O18ap))
+summ3<-modern_group %>%
+  summarize(length(modC13), mean(modC13), sd(modC13), length(modO18), mean(modO18), sd(modO18))
+M<-merge(summ,summ2)
+M[,3:8]<-round(M[3:8], 2)
+M
+summ3
+sum(M$`length(C13ap)`)
+```
+```{r}
+carbon_group<-iso %>%
+  select(lab_n:Genus, C13ap)%>%
+  filter(!(is.na(C13ap)))%>%
+  group_by(Site, Genus)
+oxygen_group<-iso %>%
+  select(lab_n:Genus, O18ap)%>%
+  filter(!(is.na(O18ap)))%>%
+  group_by(Site, Genus)
+
+carbon_group
+oxygen_group
+```
+
+```{r results='asis'}
+stargazer(attitude, header=FALSE, type='html')
+```
+
+ANOVA PRACTICE
+
+```{r}
+carbon<-lm(iso$C13ap~iso$Genus*iso$Site)
+anova(carbon)
+```
+```{r}
+oxygen<-lm(iso$O18ap~iso$Genus*iso$Site)
+anova(oxygen)
+```

### 2017-11-12 Article analysis.Rmd
DONE!

--- 2017-10-03 Article analysis.Rmd
@@
 author: "Maria C. Codlin"
-date: "September 21, 2017"
-output: html_document
+date: "November 12, 2017"
+output: 
+html_document:
+  toc: true
+  toc_depth: 4
+  toc_float: true
 ---
-###Table of contents?
 
-###Introduction to this document, the contents, and the article.
+##Introduction
+In this document, I outline the reanalyis of Somerville et al.
+(2016), who examine the evidence for captive management of leproids at Teotihuacan, one of the earliest and largest urban centers in the Americas.
+Somerville et al.
+(2016) analysed the carbon and oxygen stable isotopes from the bone apatite of archaeological specimens from two genera of lepoids, Lepus (hares), and Sylvilagus (cottontails).
+Carbon is used as a proxy for human impact on the diet of these animals: higher δ13C apatite ratios imply an larger component of the diet was made up by human cultivated C4 or CAM plants such as maize.
+They incorperate the stable carbon isotope analysis with stable oxygen isotope analysis, as δO18 apatite ratios can be used in this context as a proxy for local humidity, providing a measure of control over environmental changes that might have influence baseline signatures of δ13C.
 
-Before I run any parts of the analysis, I load all the packages I use in this file
-```{r load libraries}
-require(dplyr)
-require(stringr)
-require(ggplot2)
-require(qwraps2)
-```
-First in excel, I deleted two rows of non-table data, i.e.
-title, and a note, and saved the file (S2_Table) as a CSV in the project directory.
-Then I read it into R and converted it to a table using dplyr
-```{r readfile}
-iso<-read.csv("S2_Table.csv")
-iso<-tbl_df(iso_total)
-head(iso_total)
+Their dataset (S2_Table.csv) contains the carbon and isotope ratios from five different parts of the city, some of which overlap slightly in period of occupation.
+The Moon Pyramid, Teopancazco and Oztoyahualco sites are the three that cover the classic period (150BC-550AD) occupation of Teotihuacan, and overlap somewhat in their occupation ranges.
+Puerta 5 and Cuevas (here Tunels y Cuevas), refer respectively to the epi-classic (600-1150AD) and post-classic (Aztec, 1150-1500AD) periods of occupation at the site.
+They also analysed a sample of "modern" (1892-1966AD) comparative specimens from the Basin of Mexico (S1_Table.csv), although these were not taken from Teotihuacan valley, they should be fairly representative of the recent agricultural landscape in the area.
+
+As previous research had indicated the neighborhood of Ozotoyahualco might have specialised in raising cottontail rabbits, Somerville et al.
+(2016) hypothesised that this site would have a higher δ13C than other parts of the city.
+Similarly, due to previous research on urban provisioning and the declining prevalence of deer in the region during the classic period, they hypothesised that management of small animals such as leproids would become more important over time, and that this should be associated with an increase in δ13C across the sites during Teotihuacan's classic period occupation.
+Finally, due to certain biological and behavioural differences between Lepus and Sylvilagus, the authors hypotheses that while both may have been managed by humans, Sylvilagus would be more suitable to captive management in an urban center, and thus would likely have a higher δ13C than Lepus.
+
+###Outline of their project and my replication
+
+In their analysis, they perform a variety of descriptive and interpretive analyses, as well as vizualisation of the data. I plan to replicate all of their tests.
+
+1. **Load** and **clean** data to remove outliers and any samples that failed their quality check
+2. Test for statistical differences between genera by calculating simple **descriptive statistics** and analysis with **student's T-test**. 
+3.
+Two, **two-factor analysis of variance tests** to evaluate the effect of genus and site on δ13C and δ18O as well as potential for interaction effects between genus and site.
+They also present a summary table of means and standard deviation broken down by site and genus.
+4.
+After genus was found to have no effect, the analysis focused on site.
+They performed a **one-way ANOVA** and a **post-hoc Tukey test** on each of δ13C and δ18O to assess which sites had an effect on the two variables.
+One site was found to have a sigificant effect on δ13C, so a **Mann-Whitney-U** (two sample Wilcox rank sum test) test was employed to compare the means of that one site to the remainder of the sample.
+5.
+The final step was visualising the data in a **scatterplot** of the means and standard deviation of each site for the two main variables (δ13C and δ18O) and two **boxplots** grouped by site.
+In this step they interpret ideas about how δ13C increased over time at Teotihuacan, and use **Pearson's test** for correlation between the two variables to rule out environmental causes for the interpretations about δ13C.
+
+##Step 1. Load and clean data
+
+First I run a code to clear the workspace, to avoid too many variables building up in my environment and start with a blank slate each time I run the entire length of code. 
+```{r clear workspace}
+rm(list=ls())
 ```
 
-To make life easier, I changed the names of two key variables. First I tried 18O and 13C as the variable names, but it threw me an error, I think by starting with a number. 
-```{r }
-names(iso)[12]<-"C13ap"
-names(iso)[13]<-"O18ap"
-names(iso)
+### Load data
+
+I read in the archaeological data (S1_table).
+In excel, I deleted any rows of non-table data, i.e.
+title, and a note, and saved the file (S2_Table) as a CSV in the project directory.
+Then I read it into R and converted it to a table using dplyr.
+I trimed white space at beginnings and ends of things where there shouldn't be any, and kept strings as for now.
+Initially I used the skip code to remove the headers of the table instead of excel, but I had trouble at some point and simply removed these in the saved file.
+
+```{r reading in the archaeological data}
+require(curl)
+require(dplyr)
+f <- curl("https://raw.githubusercontent.com/mazattack/Article-analysis/master/S2_Table.csv") #This is the archaeological data
+iso_total <- read.csv(f, header = TRUE, stringsAsFactors = FALSE, strip.white=TRUE)
+iso_total<-tbl_df(iso_total)
+head(iso_total)
+str(iso_total)
 ```
-This is my first attempt at a box plot! I figured it out eventually
-```{r }
-g<-ggplot(iso, aes(x=Site, y=O18ap))
-g+geom_boxplot(aes(color=Genus))+labs(x="" ,y="δ18O apatite-VSMOW (‰)")
+I did the same with the modern data. 
+```{r readfile modern}
+
+f <- curl("https://raw.githubusercontent.com/mazattack/Article-analysis/master/S1_Table.csv") #This is the archaeological data
+modern<-read.csv(f, header=TRUE, sep = ",",strip.white=TRUE, stringsAsFactors = FALSE)
+
+modern<-tbl_df(modern)
+head(modern)
+str(modern)
 ```
-Continuation of cleaning.
-Some samples were removed from analysis due to the likely hood of contamination or diagenesis.
-" Five apatite samples with C/P ratios <0.10, three apatite samples with IR-SF values < 2.00, and five apatite samples with IR-SF values >4.00 were excluded from subsequent discussions"
 
-```{r }
-iso<-filter(iso_total, between (C.P, 0.1, 0.5)&between (IR.SF, 2.0, 4.0))
+###Clean data
+
+To make life easier, I changed the names of two key variables because they were long and involved weird characters.
+First I tried 18O and 13C as the variable names, but it threw me an error, I think by starting with a number.
+```{r rename}
+names(iso_total)[12]<-"C13ap" #I used names(iso_total) to get the position of these for renaming
+names(iso_total)[13]<-"O18ap"
+names(iso_total)[1]<-"lab_n"
+names(modern)[10]<-"modC13"
+names(modern)[11]<-"modO18"
+```
+
+The first step in their analysis is to remove samples potentially affected by diagenisis.
+In this article, they tested a random sample (n=101) of bone for potential post-depositional alteration to the mineral structure.
+Using FTIR, they tested the carbonate to phospate ratios (C/P) and infrared splitting factor (IR-SF) using FTIR and removed samples which fell outside of the expected ranges of 0.1 and 0.5 for C/P and 2.0 and 4.0 for IR-SF.
+The samples that they removed are marked in grey in the supplimentary data.
+
+Here, I used the filter function to select only those rows where C/P and IR-SF fell within the expected ranges, and saved this as a new dataset "iso".
+I also needed to make sure the rows with NA values were still included, since only a random sample was analysed.
+I did this using the | notation which functions like "OR".
+Since the FTIR analysis returned both C/P and IR-SF values, I only need to include one is.na function here.
+I also excluded an outlier with an extremely low δO18 that they excluded by converting this number to NA, as the sample's δ13C value was still incorperated in the analysis.
+
+```{r filter out diagenisis}
+#to filter out only Lepus and Sylvilagus genus, I first needed to define the column as a factor
+iso_total$Genus<-as.factor(iso_total$Genus)
+iso<-filter(iso_total, (Genus=="Lepus" | Genus=="Sylvilagus") & between (C.P, 0.1, 0.5)&between (IR.SF, 2.0, 4.0) | (is.na(IR.SF))) #brackets around the Genus filter were important here, otherwise the | worked on everything that came after
+iso$O18ap[iso$O18ap==11.9]<-NA #couldn't figure out how to get this to work with dplyr or chaining
 iso            
@@
 
-###Outline of what needs to be done. 
-Datasets: 1. Isotopes of modern specimens. 2. Isotopes of ancient specimens.
-
-##Modification of dataset prior to analysis
-- calibration of carbon isotope ratios due to atmospheric changes in the past 130 years to make the modern specimens in the dataset comparable to the ancient ones.**It is not clear whether this is incorperated in the supplimental data or not**
-
-##Statistical analyses:
-- Parametric tests: Independent samples t-test (groups of two) and ANOVA (and two or more)
--Non parametic: Mann-Whitney U test (for Oztoyahualco and pooled other areas) due to very different sample sizes.
--To see if carbon and oxygen isotopes covary, bivariate Pearson's correlation was used. Sig assumed at a=0.05.
-
-##Mixing model to approximate percentage of C4 and CAM plants in diet:
-%C4CAM= ((delta_13_C_apatite - 9.7)-(-26.1))/14.1
-Due to the level of uncertainties, these percentages are used to reflect the relative contrabutions of C4 and CAM plants between individuals rather than as absolute data.
-
-```{r summary}
-summarise(iso)
-
+There were a couple rows at the end of the dataset with no data in them so this excluded those from the dataset too. 
+```{r}
+iso<-filter(iso, !(is.na(iso$Genus))) #removes empty rows
 ```
 
-*Because of errors later on, I found I also needed to trim the space at the end of some of the Genus names.* 
-```{r trim white space}
-iso_total$Genus<-str_trim(iso_total$Genus)
-iso_total$Genus<-as.factor(iso_total$Genus)
-levels(iso_total$Genus)
-```
+##Step 2. Test for differences between genera
+
+### Descriptive statistics
+In order to test whether there was a significant difference in the isotopic values between Lepus and Sylvilagus, Somerville et al.
+(2016) first provide some summary statistics for the dataset which I calculated to ensure I had correctly removed all the outliers from the same.
+
+**Descriptive statistics from the article (Somerville et al. 2016:10)**
+
+*Archaological specimens:*
+
+  δ13Capatite = -7.8 ± 2.4‰(N = 114, 1 S.D.)
+
+  δ18Oapatite = 26.0‰(N = 113, 1 S.D.)
+
+*Modern reference specimens:*
+
+  δ13Capatite = -12.4 ± 1.8‰(N = 13, 1 S.D.)
+
+  δ18Oapatite = 26.1‰± 1.4 (N = 13, 1 S.D.)
 
@@
   list("Archaeological samples" = 
-         list("δC13 mean (sd)" = ~ mean_sd(C13ap),
-              "δO18 mean (sd)" = ~ mean_sd(O18ap)),
+        round (c("δC13 n" = sum(!is.na(iso$C13ap)),
+              "δC13 mean" =  mean(iso$C13ap, na.rm=TRUE),
+              "δC13 sd" =  sd(iso$C13ap, na.rm=TRUE),
+              "δO18 n" = sum(!is.na(iso$O18ap)),
+              "δO18 mean" =  mean(iso$O18ap, na.rm=TRUE),
+              "δO18 sd" =  sd(iso$O18ap, na.rm=TRUE)),2),
        "Modern samples" = 
-         list("δC13 mean (sd)" = ~ mean_sd(modern$modC13),
-              "δO18 mean (sd)" = ~ mean_sd(modern$modO18))
+         round(c("δC13 n" = sum(!is.na(modern$modC13)),
+              "δC13 mean" =  mean(modern$modC13, na.rm=TRUE),
+              "δC13 sd" =  sd(modern$modC13, na.rm=TRUE),
+              "δO18 n" = sum(!is.na(modern$modO18)),
+              "δO18 mean" =  mean(modern$modO18, na.rm=TRUE),
+              "δO18 sd" =  sd(modern$modO18, na.rm=TRUE)),2)
        )
 
-summary_table(iso, my_sum[1])
-summary_table(iso, my_sum[2])
+my_sum
 ```
+My summary statistics closely match those presented in the article.
+
+Note: the first time I tried this it was not a match and so I realised I needed to keep the na values in the filter above.
+The second time it was also not correct so I added two lines to count the number of values in the two columns of interest to see if the sample sizes match.
+I noted the single value from delta O18 to remove but I still have one extra row in there than their sample number.
+I checked on the excel, and there is no evidence of this extra row that they excluded.
+I took a guess that this may have been a species listed in the spreadsheet as Lepus?
+and removal of this specimen now produces averages that match the data presented in the article.
+
+###T-tests
+
+They then calculate student's T-tests to evaluate whether there is a significant difference in the means of the δ13C and δ18O between Lepus and Sylvilagus.
+
+First I created a short hand name for each variable group
+```{r}
+LC<-iso$C13ap[iso$Genus=="Lepus"]
+SC<-iso$C13ap[iso$Genus=="Sylvilagus"]
+LO<-iso$O18ap[iso$Genus=="Lepus"]
+SO<-iso$O18ap[iso$Genus=="Sylvilagus"]
+```
+
+####Carbon apatite two sample T-test
+```{r}
+t.test(LC, SC, var.equal =TRUE, alternative = "two.sided")
+```
+
+####Oxygen apatite two sample T-test
+```{r}
+t.test(SO, LO, var.equal =TRUE, alternative = "two.sided")
+```
+**Summary of my results**
+
+Carbon apatite t-sample t-test comparing Lepus and Sylvilagus: t = 0.5185, df = 112, p-value = 0.6051
+
+Oxygen apatite t-sample t-test comparing Lepus and Sylvilagus: t = -0.83133, df = 111, p-value = 0.4076
+
+**Test values reported in Somerville et al. (2016:10)**
+
+δ13Capatite values (t[110] = -0.806, p = 0.422)
+
+δ18Oapatite values (t[110] = 0.428, p = 0.669)
+
+Both of these tests returned different test statistics, degrees of freedom and p-values to those reported by Somerville et al.
+(2016).
+I tried multiple variations of the t-test in R, from Welsh t-test and one sided t-tests to one sided Welsh t-tests and none of the tests produced a similar result to those reported by Somerville.
+It later became clear there was an error in either their reported data or supplimentary data - see step 3 for explanation.
+That said, the outcome of all tests is the same, there is no significant difference in the means of the samples.
+
+###Test for normality
+While they did not report any tests for normality, I also decided to do a cursory check of the suitability of students T-test to these data.
+I check that the data fits a normal distribution by viewing their histogram, qnorm and qqline plots as well as performing the shapiro test for normality.
+Finally, I also test for equal variance between the samples.
+
+####δ13C
+
+```{r carbon normality test}
+require(ggplot2)
+a_hist<-ggplot(iso, aes(C13ap))
+a_hist+geom_histogram(bins= 15)+ facet_wrap(~Genus, scales="free") #I couldn't figure out how to get rid of the blank chart at the start
+qqnorm(LC,xlab="Theoretical",main="Lepus")
+qqline(LC)
+
+qqnorm(SC,xlab="Theoretical",main="Sylvilagus")
+qqline(SC)
+
+shapiro.test(LC) #Lepus
+shapiro.test(SC) #Sylvilagus
+
+var.test(LC, SC) #test of equal variance
+```
+In these tests, Sylvilagus looks approximately normal while Lepus has a more noticeable skew.
+However both pass the Shapiro-Wilk test and have equal variance (although would have failed this test if alpha were 0.1) so are likely normal enough to trust the t-test results.
+
+####δ18O
+
+```{r oxygen normality test}
+require(ggplot2)
+a_hist<-ggplot(iso, aes(O18ap))
+a_hist+geom_histogram(bins=15)+ facet_wrap(~Genus, scales="free")
+
+qqnorm(LO,xlab="Theoretical",main="Lepus")
+qqline(LO)
+
+qqnorm(SO,xlab="Theoretical",main="Sylvilagus")
+qqline(SO)
+
+shapiro.test(LO) #Lepus
+shapiro.test(SO) #Sylvilagus
+
+var.test(LO, SO)
+```
+
+Here, Lepus fits the normal distrbution well, while Sylvilagus is divergent.
+Looking at the histgram again, it could potentitally be bimodal.
+If this were my data, I would explore that possibility further as it could potentially indicate sylvilagus a group of rabbits with different environmental backgrounds.
+However, both past the Shapiro-Wilk test and appear to have equal variances so the t-test should be ok.
+
+##Step 3: Test for effects of site and genus 
+The article uses **two-factor analysis of variance tests** to evaluate the effect of genus and site on δ13C and δ18O as well as potential for interaction effects between genus and site.
+They also present a summary table of means and standard deviation broken down by site and genus.
+
+###Summary table:
+I created the summary table to compare to their Table 2
+
+First I factorize Genus and define levels of the modern dataset in order to merge with the archaeological dataset
+```{r create genus column in modern}
+Genus<- as.factor(modern$Species) 
+modern<-cbind(modern, Genus)
+levels(modern$Genus)[levels(modern$Genus)=="Lepus californicus texianus"]<-"Lepus"
+levels(modern$Genus)[levels(modern$Genus)=="Sylvilagus floridanus"]<-"Sylvilagus"
+levels(modern$Genus)[levels(modern$Genus)=="Sylvilagus floridanus orizabae"]<-"Sylvilagus"
+levels(modern$Genus)    
+```
+Here I code the merge, selecting only the necessary columns
+```{r merge modern and iso datasets}
+new<-select(iso, lab_n, Site, Phase, Genus, C13ap, O18ap)
+new2<-select(modern, lab_n=Lab_num, Site=State, Phase=Collection.date..yrs.AD., Genus, C13ap=modC13, O18ap=modO18)%>%mutate(Site="modern") #Giving the modern dataset a "site" variable
+iso2<-rbind(filter(new,(!(is.na(C13ap)))), new2) #merging with rbind
+iso2
+```
+Finally I put together the summary table. I decided not to also include the subtotal row in this table, as that would be needlessly complicated and is not something I would typically use R for.
+
+```{r constructing the summary table}
+iso2$Genus <-as.factor(iso2$Genus)
+iso2_g <- group_by(iso2, Site, Genus)
+sum_tot<- summarize(iso2_g, length(C13ap), mean(C13ap), sd(C13ap), length(O18ap), mean(O18ap), sd(O18ap))
+
+tot_tot<-iso2%>%
+  group_by(Genus)%>%
+  summarize(length(C13ap), mean(C13ap), sd(C13ap), length(O18ap), mean(O18ap), sd(O18ap))
+tot_tot<-mutate(tot_tot, Site="Total")
+tot<-rbind.data.frame(sum_tot, tot_tot)
+tot
+```
+
+Note the differences across these summary statistics compared to the table provided in the article (Table 2). 
+<img src="img/Orig_sum_table.jpg" width="1000px"/>
+
+This reflects a difference in the data from the article compared to what they provided in the supplimentary data.
+I checked this in excel to confirm.
+It seems a few cases were moved between contexts and genus variables, so while this difference is not enough to substantially throw off my results, most of the tests do not match up exactly.
+
+###Two-way ANOVA
+
+The article employed two, **two-factor analysis of variance tests** to evaluate the effect of genus and site on δ13C and δ18O as well as potential for interaction effects between genus and site.
+I tested the type II anova, which would be better for the unbalanced design, however they gave almost identical p-values so I continued with type I for ease here.
+```{r carbon anova}
+carbon<-lm(iso$C13ap~iso$Genus*iso$Site)
+anova(carbon)#not sure why but moon pyramid is missing from all these ANOVA, and then we had that class and now I know.
+plot(carbon$residuals)# I also decided to check out the residuals
+```
+```{r oxygen anova}
+oxygen<-lm(iso$O18ap~iso$Genus*iso$Site)
+anova(oxygen)
+plot(oxygen$residuals)# I also decided to check out the residuals
+```
+
+**Somerville et al's (2016) two-way ANOVA results:**
+
+"For δ13C apatite, the main effect for genus was not significant (F[1,102] = 3.361), p = 0.070), but the main effect of site context was significant (F[4,102] = 5.365, p = 0.001).
+Importantly, the interaction between genus and site was not significant for δ13Capatite (F[4,102] = 0.590, p<0.670) or for δ18Oapatite (F[4,102] = 0.834, p<0.507)."
+
+While the values do not match exactly due to the error in the dataset, my two way Anova summaries match the conclusions they derived in the article.
+That Genus had no significant effect on either δ13C or δO18 apatite ratios.
+However there is a significant effect of Site location on these variables.
+Also, the residuals show no clear patterns that would indicate another major influencing variable on these data.
+
+##Step 4. Identifying the site factors effecting the isotopic signature
+
+After genus was found to have no effect, the analysis focused on site.
+Somerville et al.
+(2016) performed a **one-way ANOVA** and a **post-hoc Tukey test** on each of δ13C and δ18O to assess which sites had an effect on the two variables.
+One site was found to have a sigificant effect on δ13C, so a **Mann-Whitney-U** (two sample Wilcox rank sum test) test was employed to compare the means of that one site to the remainder of the sample.
+
+###One way ANOVA tests and post-hoc Tukey test
+####δ13C
+```{r carbon one way}
+carbon2<-lm(iso$C13ap~iso$Site) 
+m <- aov(iso$C13ap~iso$Site)
+summary(m)
+posthoc <- TukeyHSD(m, "iso$Site", conf.level = 0.95)
+posthoc 
+```
+The one way ANOVA test for δ13C also identified a significant effect of site on δ13C and the post-hoc Tukey test highlights significant differences in the δ13C between Oztoyahualco-Moon Pyramid, Teopancazco-Moon Pyramid and Túneles y Cuevas-Oztoyahualco.
+This mirrors Somerville et al.'s (2016) findings and implys that leproids at Oztoyahualco had a higher levels of C4 plants in the diet.
+This supports the theory that residents at Oztoyahualco may have been raising leproids.
+
+*Somerville et al.'s (2016:10) One-way ANOVA and Tukey results:* "Significant differences were found in δ13Capatite (F[4, 107] = 5.8, p<0.001).
+Post-hoc Tukey tests indicate that Oztoyahualco leporids had significantly higher δ13Capatite values than those of the Moon Pyramid and the Cuevas"
+
+####δ18O
+
+```{r oxygen one way}
+m <- aov(iso$O18ap~iso$Site)
+summary(m)
+posthoc <- TukeyHSD(m, "iso$Site", conf.level = 0.95)
+posthoc  
+```
+The one way ANOVA test for δ18O also identified a significant effect of site on δ18O and the post-hoc Tukey test highlights significant differences in the δ18O between Puerta 5-Moon Pyramid, Teopancazco-Oztoyahualco and Teopancazco-Puerta 5.
+This mirrors Somerville et al.'s (2016) findings.
+
+**Somerville et al.'s (2016:11) One-way ANOVA and Tukey results:**
+"Significant differences were also found in δ18Oapatite means between site contexts (F[5, 120] = 4.704, p<0.001), but no obvious temporal trends were observed (Fig 6).
+Post-hoc Tukey tests revealed significant differences in δ18O apatite between Oztoyahualco and Teopancazco (p = 0.046), between Puerta 5 and Teopancazco (p<0.001), and between Puerta 5 and the Moon Pyramid (p = 0.008)."
+
+###Mann-Whitney-U test
+Finally, to confirm that the mean δ13C samples from Oztoyahualco are significantly different from the remainder of the site, they performed a Mann-Whitney-U test.
+
+```{r Mann-Whitney U carbon}
+group1<-(iso2) %>%
+  filter(Site!="modern", Site!="Oztoyahualco") %>%
+  select(C13ap)
+group2<-(iso2) %>%
+  filter(Site=="Oztoyahualco") %>%
+  select(C13ap)
+group1<-as.numeric(unlist(group1)) #there was some weird stuff going on, but unisting it worked
+group2<-as.numeric(unlist(group2))
+wilcox.test(group1, group2)
+```
+As they expected, mean δ13C is siginifancly higher than other areas at Teotihuacan which supports the theory that residents at Oztoyahualco may have been raising leproids.
+
+**Somerville et al.'s (2016:11) Mann-Whitney-U test for δ13C apatite:** (U = 329, p = 0.001)
+
+```{r Mann-Whitney U oxygen}
+group1<-(iso2) %>%
+  filter(Site!="modern", Site!="Oztoyahualco") %>%
+  select(O18ap)
+group2<-(iso2) %>%
+  filter(Site=="Oztoyahualco") %>%
+  select(O18ap)
+group1<-as.numeric(unlist(group1))
+group2<-as.numeric(unlist(group2))
+wilcox.test(group1, group2)
+```
+No significant differences were identified for the δ18O between Oztoyahualco and the remainder of the site.
+They do not interpret this result, but I imagine it supports the conclusion they make later which is that differences in δ13C cannot be attributed to environmental change as measured by δ18O values.
+This result closely matches their test.
+
+**Somerville et al.'s (2016:11) Mann-Whitney-U test for δ18O apatite:**(U = 636, p = 0.434).
+
+##Step 5: Data visualisation
+
+The final step was visualising the data in a **scatterplot** of the means and standard deviation of each site for the two main variables (δ13C and δ18O) and two **boxplots** grouped by site.
+In this step they interpret ideas about how δ13C increased over time at Teotihuacan, and use **Pearson's test** for correlation between the two variables to rule out environmental causes for the interpretations about δ13C.
+
+###Scatterplot
+First I relabel the site factors to include the temporal information
+```{r relabel levels of Site, note that runing this twice will erase the data}
+iso2$Site = factor(iso2$Site, c("Moon Pyramid", "Teopancazco", "Oztoyahualco", "Puerta 5", "Túneles y Cuevas", "modern"), labels=c("Moon Pyramid (150 BC-AD 450)", "Teopancazco (AD 200-550)", "Oztoyahualco (AD 350-550)", "Puerta 5 (AD 600-1150)", "Túneles y Cuevas (AD 1150-1500)", "modern (AD 1892-1966)"))
+```
+Then calculate summary statistics to create the summary graph Figure 4. 
+```{r carbon v oxygen scatterplot}
+gm <- iso2 %>% 
+        group_by(Site) %>% 
+        summarise(mean_C13ap = mean(C13ap),
+                  mean_O18ap  = mean(O18ap, na.rm=T),
+                  C13SD=sd(C13ap),
+                  O18SD=sd(O18ap, na.rm=T))
+gm
+require(ggplot2)
+g<-ggplot(gm, aes(x=mean_C13ap, y=mean_O18ap, shape = Site))
+g+geom_point(data=gm, size=4)+ xlim(-16, -1) +ylim(21, 31)+
+      geom_errorbarh(data = gm,aes(xmin = mean_C13ap - C13SD,xmax = mean_C13ap + C13SD,y = mean_O18ap)) + 
+    geom_errorbar(data = gm,aes(ymin = mean_O18ap - O18SD,ymax = mean_O18ap + O18SD,x = mean_C13ap)) +
+  labs(x="δ13C apatite-VPDB (‰)", y="δ18O apatite-VSMOW (‰)") +
+  annotate(geom = "text", x = -12, y = 22, label = "note error bars are SD")
+```
+
+Compare this to their figure 4 (below)
+
+<img src="img/Orig_scatter_plot.jpg" width="1000px"/>
+
+###Box plots
+#### δ13C
+```{r carbon boxplot}
+require(stringr)
+g<-ggplot(iso2, aes(x=Site, y=C13ap))
+g<-g+geom_boxplot(aes(fill=Genus))+
+  ylim(-16, -2) +
+  labs(x="" ,y="δ13C apatite-VPDB (‰)") +
+  scale_color_manual(values=c("black", "black"))+
+  scale_fill_manual(values=c( "grey90", "grey45"))+
+  theme(axis.text.x = element_text(size=8, angle=45, hjust = 1))+
+  scale_x_discrete(labels = function(x) str_wrap(x, width = 15))+#this wraps the text on the x axis
+   coord_fixed(.45)+
+    geom_vline(aes(xintercept = 3.5), linetype=2) #this line separates classic period of Teotihuacan from epi and post Classic phases
+g
+
+ggsave(filename = "carbon_box.png", g,
+       width = 8, height = 9, dpi = 300, units = "in", device='png')
+```
+There are clear temporal trends in the δ13C values at Teotihuacan, generally increasing in the Classic period and decreasing after the classic period. 
+Compare to Figure 5 (below)
+<img src="img/Orig_carbon_plot.jpg" width="1000px"/>
+
+####δ18O
+```{r oxygen boxplot}
+require(stringr)
+g<-ggplot(iso2, aes(x=Site, y=O18ap))
+g<-g+geom_boxplot(aes(fill=Genus))+
+  ylim(18, 32) +
+  labs(x="" ,y="δ18O apatite-VSMOW (‰)") +
+  scale_color_manual(values=c("black", "black"))+
+  scale_fill_manual(values=c( "grey90", "grey45"))+
+  theme(axis.text.x = element_text(size=8, angle=45, hjust = 1))+
+  scale_x_discrete(labels = function(x) str_wrap(x, width = 15))+
+   coord_fixed(.45)+
+   geom_vline(aes(xintercept = 3.5), linetype=2)
+g
+
+ggsave(filename = "oxygen_box.png", g,
+       width = 8, height = 9, dpi = 300, units = "in", device='png')
+```
+There is no obvious temporal trend in oxygen isotopes. 
+Compare to Figure 6 (below)
+<img src="img/Orig_oxygen_plot.jpg" width="1000px"/>
+
+The observations from the previous plots suggest no major correlation between Oxygen and Carbon isotope.
+This was tested using Pearson's correlation test and the results supported their observation of no significant correlation.
+
+### Pearson's Correlation
+```{r}
+cor.test(x=iso2$C13ap, y=iso2$O18ap, method= "pearson")
+```
+**Somerville et al.'s (2016:14) Pearson's correlation:** (R = -0.003, p = 0.971).

### 2017-11-13 Article analysis notebook.Rmd
finals

@@
   pdf_document: default
+    toc:true
+	  toc_depth:4
+	  toc_float:true
+
 ---
 ###Table of contents?
+
+#Introduction
+In this document, I outline the reanalyis of Somerville et al.
+(2016), who examine the evidence for captive management of leproids at Teotihuacan, one of the earliest and largest urban centers in the Americas.
+Somerville et al.
+(2016) analysed the carbon and oxygen stable isotopes from the bone apatite of archaeological specimens from two genera of lepoids, Lepus (hares), and Sylvilagus (cottontails).
+Carbon is used as a proxy for human impact on the diet of these animals: higher δ13C apatite ratios imply an larger component of the diet was made up by human cultivated C4 or CAM plants such as maize.
+They incorperate the stable carbon isotope analysis with stable oxygen isotope analysis, as δO18 apatite ratios can be used in this context as a proxy for local humidity, providing a measure of control over environmental changes that might have influence baseline signatures of δ13C.
+
+Their dataset (S2) contains the carbon and isotope ratios from five different parts of the city, some of which overlap slightly in period of occupation.
+The Moon Pyramid, Teopancazco and Oztoyahualco sites are the three that cover the classic period (150BC-550AD) occupation of Teotihuacan, and overlap somewhat in their occupation ranges.
+Puerta 5 and Cuevas (here Tunels y Cuevas), refer respectively to the epi-classic (600-1150AD) and post-classic (Aztec, 1150-1500AD) periods of occupation at the site.
+They also analysed a sample of "modern" (1892-1966AD) comparative specimens from the Basin of Mexico (S1), although these were not taken from Teotihuacan valley, they should be fairly representative of the recent agricultural landscape in the area.
+
+As previous research had indicated the neighborhood of Ozotoyahualco might have specialised in raising cottontail rabbits, Somerville et al.
+(2016) hypothesised that this site would have a higher δ13C than other parts of the city.
+Similarly, due to previous research on urban provisioning and the declining prevalence of deer in the region during the classic period, they hypothesised that management of small animals such as leproids would become more important over time, and that this should be associated with an increase in δ13C across the sites during Teotihuacan's classic period occupation.
+Finally, due to certain biological and behavioural differences between Lepus and Sylvilagus, the authors hypotheses that while both may have been managed by humans, Sylvilagus would be more suitable to captive management in an urban center, and thus would likely have a higher δ13C than Lepus.
+
+###Outline of their project and my replication
+
+In their analysis, they perform a variety of descriptive and interpretive analyses, as well as vizualisation of the data. I plan to replicate all of their tests.
+
+1. Load and clean data to remove outliers and any samples that failed their quality check
+2. Test for statistical differences between genera by calculating simple descriptive statistics and analysis with students T-test. 
+3. Two analysis of variance tests to evaluate the effect of genus and site on δ13C and δ18O as well as potential for interaction effects between genus and site.
+4.
+After genus was found to have no effect, the analysis focused on site.
+They performed a one-way ANOVA and a post-hoc Tukey test on each of δ13C and δ18O to assess which sites had an effect on the two variables.
+One site was found to have a sigificant effect on δ13C, so a Mann-Whitney-U (two sample Wilcox rank sum test) test was employed to compare the means of that one site to the remainder of the sample.
+5.
+The final step was visualising the data in a scatterplot of the means and standard deviation of each site for the two main variables (δ13C and δ18O) and two boxplots grouped by site.
+In this step they interpret ideas about how δ13C increased over time at Teotihuacan, and use Pearson's test for correlation between the two variables to rule out environmental causes for the interpretations about δ13C.
 
@@
 
-To make life easier, I changed the names of two key variables. First I tried 18O and 13C as the variable names, but it threw me an error, I think by starting with a number. 
+To make life easier, I changed the names of two key variables because they were long and involved weird characters.
+First I tried 18O and 13C as the variable names, but it threw me an error, I think by starting with a number.
 ```{r rename}
@@
 
-In this article, they tested a random sample (n=101) of bone for potential post-depositional alteration to the mineral structure.Using FTIR, they tested the carbonate to phospate ratios (C/P) and infrared splitting factor (IR-SF) using FTIR and removed samples which fell outside of the expected ranges of 0.1 and 0.5 for C/P and 2.0 and 4.0 for IR-SF.
+In this article, they tested a random sample (n=101) of bone for potential post-depositional alteration to the mineral structure.
+Using FTIR, they tested the carbonate to phospate ratios (C/P) and infrared splitting factor (IR-SF) using FTIR and removed samples which fell outside of the expected ranges of 0.1 and 0.5 for C/P and 2.0 and 4.0 for IR-SF.
 The samples that they removed are marked in grey in the supplimentary data.
@@
 Since the FTIR analysis returned both C/P and IR-SF values, I only need to include one is.na function here.
-I also excluded an outlier with an extremely low delta O18 that they excluded by converting this number to NA, as the samples delta C13 value was still incorperated in the analysis.
+I also excluded an outlier with an extremely low δO18 that they excluded by converting this number to NA, as the sample's δ13C value was still incorperated in the analysis.
 ```{r filter out diagenisis}
@@
 ```
+There were a couple rows at the end of the dataset with no data in them so this excluded those from the dataset too. 
 ```{r}
@@
 
-First, I will check that the data fits a normal distribution by comparing it to the qnorm and qqline plots as well as histogram.
+First, I check that the data fits a normal distribution by viewing their histogram,  qnorm and qqline plots as well as performing the shapiro test for normality.
 
@@
 a_hist<-ggplot(iso, aes(C13ap))
-a_hist+geom_histogram(bins= 15)+ facet_wrap(~Genus, scales="free")
+a_hist+geom_histogram(bins= 15)+ facet_wrap(~Genus, scales="free") #I couldn't figure out how to get rid of the blank chart at the start
 ```
@@
 ```{r qqplots carbon}
-LC<-iso$C13ap[iso$Genus=="Lepus"]
-SC<-iso$C13ap[iso$Genus=="Sylvilagus"]
+
 qqnorm(LC,xlab="Theoretical",main="Lepus")
@@
 
-  shapiro.test(LC)
-  shapiro.test(SC)
+  shapiro.test(LC) #Lepus
+  shapiro.test(SC) #Sylvilagus
 ```
@@
 ```{r qqplots Oxygen}
-LO<-iso$O18ap[iso$Genus=="Lepus"]
-SO<-iso$O18ap[iso$Genus=="Sylvilagus"]
+
 qqnorm(LO,xlab="Theoretical",main="Lepus")
@@
 I tried multiple variations of the t-test in R, from Welsh t-test and one sided t-tests to one sided Welsh t-tests and none of the tests produced a similar result to those reported by Somerville.
-In fact, the two sample t-test for equal varience that I ran for both carbon and oxygen data matches most closely with the result reported for the opposite isotopic data.
-i.e.
-my oxygen looks most similar to their carbon.
-I have a number of hypothesis for why these tests may vary: the SPSS test is vastly different to the R t-tests, they are excluding samples from the test and not reporting it, as indicated by the difference in df between our tests, or there was an error in their reporting of the data.
+It later became clear there was an error in either their reported data or supplimentary data - see step 3 for explanation.
 That said, the outcome of all tests is the same, there is no significant difference in the means of the samples.
-I later realised that the data they provide is slightly differerent from the results presented and this is likely the reason for the differences observed in this test.
-This is most clearly observed in the table below and comparison to their table 2.
+
+I also conducted tests of normality to assess the suitability of students T-test to these data.
 
@@
 iso2_g <- group_by(iso2, Site, Genus)
-sum_tot<- summarize(iso2_g, length(C13ap), mean(C13ap), sd(C13ap), length(O18ap), mean(O18ap), sd(O18ap))
+sum_tot<- iso2_g%>%
+  summarize(length(C13ap), mean(C13ap), sd(C13ap), length(O18ap), mean(O18ap), sd(O18ap))
 
@@
 carbon<-lm(iso$C13ap~iso$Genus*iso$Site)
-anova(carbon) #not sure why but moon pyramid is missing from all these ANOVA
+anova(carbon)#not sure why but moon pyramid is missing from all these ANOVA
+plot(carbon$residuals)
 ```
@@
 ```{r}
-carbon<-lm(iso$C13ap~iso$Site)
-summary(carbon)
+carbon2<-lm(iso$C13ap~iso$Site)
+anova(carbon2)
+m <- aov(iso$C13ap~iso$Site)
+summary(m)
+posthoc <- TukeyHSD(m, "iso$Site", conf.level = 0.95)
+posthoc
+
 ```
+```{r Mann-Whitney U carbon}
+group1<-(iso2) %>%
+  filter(Site!="modern", Site!="Oztoyahualco") %>%
+  select(C13ap)
+group2<-(iso2) %>%
+  filter(Site=="Oztoyahualco") %>%
+  select(C13ap)
+group1<-as.numeric(unlist(group1))
+group2<-as.numeric(unlist(group2))
+wilcox.test(group1, group2)
+```
+```{r Mann-Whitney U oxygen}
+group1<-(iso2) %>%
+  filter(Site!="modern", Site!="Oztoyahualco") %>%
+  select(O18ap)
+group2<-(iso2) %>%
+  filter(Site=="Oztoyahualco") %>%
+  select(O18ap)
+group1<-as.numeric(unlist(group1))
+group2<-as.numeric(unlist(group2))
+wilcox.test(group1, group2)
+```
+```{r}
 
-```{r}
-oxygen<-lm(iso$O18ap~iso$Site)
-summary(oxygen)
 ```
@@
 
+```
+
+```{r}
+m <- aov(iso$O18ap~iso$Site)
+summary(m)
+posthoc <- TukeyHSD(m, "iso$Site", conf.level = 0.95)
+posthoc  
 ```
@@
 contextual comparisons thus combine Sylvilagus and Lepus specimens*
+
+#Visualing the data in graphs
+
+gm <- iso2 %>% 
+        group_by(Site) %>% 
+        summarise(mean_C13ap = mean(C13ap),
+                  mean_O18ap  = mean(O18ap, na.rm=T),
+                  C13SE=sqrt(var(C13ap))/length(C13ap),
+                  O18SE=sqrt(var(O18ap, na.rm=T))/length(O18ap))
+
+```{r relabel levels of Site, note that runing this twice will erase the data}
+iso2$Site = factor(iso2$Site, c("Moon Pyramid", "Teopancazco", "Oztoyahualco", "Puerta 5", "Túneles y Cuevas", "modern"), labels=c("Moon Pyramid (150 BC-AD 450)", "Teopancazco (AD 200-550)", "Oztoyahualco (AD 350-550)", "Puerta 5 (AD 600-1150)", "Túneles y Cuevas (AD 1150-1500)", "modern (AD 1892-1966)"))
+```
+```{r carbon v oxygen scatterplot}
+gm <- iso2 %>% 
+        group_by(Site) %>% 
+        summarise(mean_C13ap = mean(C13ap),
+                  mean_O18ap  = mean(O18ap, na.rm=T),
+                  C13SD=sd(C13ap),
+                  O18SD=sd(O18ap, na.rm=T))
+gm
+
+g<-ggplot(gm, aes(x=mean_C13ap, y=mean_O18ap, shape = Site))
+g+geom_point(data=gm, size=4)+ xlim(-16, -1) +ylim(21, 31)+
+      geom_errorbarh(data = gm,aes(xmin = mean_C13ap - C13SD,xmax = mean_C13ap + C13SD,y = mean_O18ap)) + 
+    geom_errorbar(data = gm,aes(ymin = mean_O18ap - O18SD,ymax = mean_O18ap + O18SD,x = mean_C13ap)) +
+  labs(x="δ13C apatite-VPDB (‰)", y="δ18O apatite-VSMOW (‰)") +
+  annotate(geom = "text", x = -12, y = 22, label = "note error bars are SD")
+
+```
+
+```{r carbon boxplot}
+
+g<-ggplot(iso2, aes(x=Site, y=C13ap))
+g<-g+geom_boxplot(aes(fill=Genus))+
+  ylim(-16, -2) +
+  labs(x="" ,y="δ13C apatite-VPDB (‰)") +
+  scale_color_manual(values=c("black", "black"))+
+  scale_fill_manual(values=c( "grey90", "grey45"))+
+  theme(axis.text.x = element_text(size=8, angle=45, hjust = 1))+
+  scale_x_discrete(labels = function(x) str_wrap(x, width = 15))+
+   coord_fixed(.45)+
+    geom_vline(aes(xintercept = 3.5), linetype=2)
+g
+
+ggsave(filename = "carbon_box.png", g,
+       width = 8, height = 9, dpi = 300, units = "in", device='png')
+```
+
+```{r oxygen boxplot}
+
+g<-ggplot(iso2, aes(x=Site, y=O18ap))
+g<-g+geom_boxplot(aes(fill=Genus))+
+  ylim(18, 32) +
+  labs(x="" ,y="δ18O apatite-VSMOW (‰)") +
+  scale_color_manual(values=c("black", "black"))+
+  scale_fill_manual(values=c( "grey90", "grey45"))+
+  theme(axis.text.x = element_text(size=8, angle=45, hjust = 1))+
+  scale_x_discrete(labels = function(x) str_wrap(x, width = 15))+
+   coord_fixed(.45)+
+   geom_vline(aes(xintercept = 3.5), linetype=2)
+g
+
+ggsave(filename = "oxygen_box.png", g,
+       width = 8, height = 9, dpi = 300, units = "in", device='png')
+```
+
+```{r}
+library(Hmisc)
+cor.test(x=iso2$C13ap, y=iso2$O18ap, method= "pearson")
+```
+
+---
+title: "Article analysis"
+author: "Maria C. Codlin"
+date: "September 21, 2017"
+output: html_document
+---
+###Table of contents?
+
+###Introduction to this document, the contents, and the article.
+
+Before I run any parts of the analysis, I load all the packages I use in this file
+```{r load libraries}
+require(dplyr)
+require(stringr)
+require(ggplot2)
+require(qwraps2)
+```
+First in excel, I deleted two rows of non-table data, i.e.
+title, and a note, and saved the file (S2_Table) as a CSV in the project directory.
+Then I read it into R and converted it to a table using dplyr
+```{r readfile}
+iso<-read.csv("S2_Table.csv")
+iso<-tbl_df(iso_total)
+head(iso_total)
+```
+
+To make life easier, I changed the names of two key variables. First I tried 18O and 13C as the variable names, but it threw me an error, I think by starting with a number. 
+```{r }
+names(iso)[12]<-"C13ap"
+names(iso)[13]<-"O18ap"
+names(iso)
+```
+This is my first attempt at a box plot! I figured it out eventually
+```{r }
+g<-ggplot(iso, aes(x=Site, y=O18ap))
+g+geom_boxplot(aes(color=Genus))+labs(x="" ,y="δ18O apatite-VSMOW (‰)")
+```
+Continuation of cleaning.
+Some samples were removed from analysis due to the likely hood of contamination or diagenesis.
+" Five apatite samples with C/P ratios <0.10, three apatite samples with IR-SF values < 2.00, and five apatite samples with IR-SF values >4.00 were excluded from subsequent discussions"
+
+```{r }
+iso<-filter(iso_total, between (C.P, 0.1, 0.5)&between (IR.SF, 2.0, 4.0))
+iso            
+```
+
+###Outline of what needs to be done. 
+Datasets: 1. Isotopes of modern specimens. 2. Isotopes of ancient specimens.
+
+##Modification of dataset prior to analysis
+- calibration of carbon isotope ratios due to atmospheric changes in the past 130 years to make the modern specimens in the dataset comparable to the ancient ones.**It is not clear whether this is incorperated in the supplimental data or not**
+
+##Statistical analyses:
+- Parametric tests: Independent samples t-test (groups of two) and ANOVA (and two or more)
+-Non parametic: Mann-Whitney U test (for Oztoyahualco and pooled other areas) due to very different sample sizes.
+-To see if carbon and oxygen isotopes covary, bivariate Pearson's correlation was used. Sig assumed at a=0.05.
+
+##Mixing model to approximate percentage of C4 and CAM plants in diet:
+%C4CAM= ((delta_13_C_apatite - 9.7)-(-26.1))/14.1
+Due to the level of uncertainties, these percentages are used to reflect the relative contrabutions of C4 and CAM plants between individuals rather than as absolute data.
+
+```{r summary}
+summarise(iso)
+
+```
+
+*Because of errors later on, I found I also needed to trim the space at the end of some of the Genus names.* 
+```{r trim white space}
+iso_total$Genus<-str_trim(iso_total$Genus)
+iso_total$Genus<-as.factor(iso_total$Genus)
+levels(iso_total$Genus)
+```
+
+```{r sum_table}
+my_sum <- 
+  list("Archaeological samples" = 
+         list("δC13 mean (sd)" = ~ mean_sd(C13ap),
+              "δO18 mean (sd)" = ~ mean_sd(O18ap)),
+       "Modern samples" = 
+         list("δC13 mean (sd)" = ~ mean_sd(modern$modC13),
+              "δO18 mean (sd)" = ~ mean_sd(modern$modO18))
+              )
+
+summary_table(iso, my_sum[1])
+summary_table(iso, my_sum[2])
+```
+
+```{r}
+stargazer(iso_total$Element, align = TRUE)
+
+```
+
+```{r summary}
+sum(!is.na(iso$C13ap))#the ! means it will return the number that is not NA
+sum(!is.na(iso$O18ap))
+arch_sum<-summarize(iso, meanC13=mean(C13ap, na.rm=TRUE), meanO18=mean(O18ap, na.rm=TRUE), sdC13=sd(C13ap, na.rm=TRUE), sdO18=sd(O18ap, na.rm=TRUE))%>%print()
+
+mod_sum<-summarize(modern, meanC13=mean(modC13, na.rm=TRUE), meanO18=mean(modO18, na.rm=TRUE), seC13=std.error(modC13, na.rm=TRUE), sdO18=sd(modO18, na.rm=TRUE))%>%print()
+
+```
+
+These results also differ from what was presented. So I will test to see if they removed outliers prior to t-test. I will use a box plot to identify outliers
+```{r}
+boxplot(SC, ylim = c(), main = "dC13 ap", xlab = "Sylvilagus")
+boxplot(LC, ylim = c(), main = "dC13 ap", xlab = "Lepus")
+boxplot(SO, ylim = c(), main = "dO18 ap", xlab = "Sylvilagus")
+boxplot(LO, ylim = c(), main = "dO18 ap", xlab = "Lepus")
+```
+
+*Working out the summary table*
+```{r group by genus}
+modern_group<-modern %>%
+  select(modC13:Genus)%>%
+  group_by(Genus)
+```
+
+```{R}
+summ<-carbon_group%>%
+  summarize(length(C13ap), mean(C13ap), sd(C13ap))
+summ2<-oxygen_group %>%
+  summarize(length(O18ap), mean(O18ap), sd(O18ap))
+summ3<-modern_group %>%
+  summarize(length(modC13), mean(modC13), sd(modC13), length(modO18), mean(modO18), sd(modO18))
+M<-merge(summ,summ2)
+M[,3:8]<-round(M[3:8], 2)
+M
+summ3
+sum(M$`length(C13ap)`)
+```
+```{r}
+carbon_group<-iso %>%
+  select(lab_n:Genus, C13ap)%>%
+  filter(!(is.na(C13ap)))%>%
+  group_by(Site, Genus)
+oxygen_group<-iso %>%
+  select(lab_n:Genus, O18ap)%>%
+  filter(!(is.na(O18ap)))%>%
+  group_by(Site, Genus)
+
+carbon_group
+oxygen_group
+```
+
+```{r results='asis'}
+stargazer(attitude, header=FALSE, type='html')
+```
+
+ANOVA PRACTICE
+
+```{r}
+carbon<-lm(iso$C13ap~iso$Genus*iso$Site)
+anova(carbon)
+```
+```{r}
+oxygen<-lm(iso$O18ap~iso$Genus*iso$Site)
+anova(oxygen)
+```
